Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework

### 2023-08-15 manuscript/manuscript_word.Rmd
1st round revision done based on Dietz's comments.

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product. Here we propose the following five basic measures to put the Triple P framework into practice: 1) the acknowledgement of the contribution and limitations of actualistic experiments properly; 2) the normalization the ethological and ethnographic data collection in experimental projects; 3) the involvement of avocational as well as novice participants; 4) the collaboration across labs on a global scale; and 5) the development of an open-access repository for data reuse.  \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

\vspace{12pt}

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology, which tends to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], the Triple P framework aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, we advocate the following five measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) normalizing the ethological and ethnographic data collection in experimental projects; 3) encouraging the involvements of avocational as well as novice participants; 4) boosting the collaboration across labs on a global scale; 5) building an open-access repository for data reuse. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

In the past decades, experimental archaeology has witnessed growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. There is no doubt that controlled experiments conducted on stone artifacts [@li2022], particularly those regarding fracture mechanics [@cotterell1992], provide some foundational and irreplaceable insights into our understanding of the role of lithic technology in prehistory, and unequivocally this line of inquiry should be celebrated and promoted to carry on. Nonetheless, it is oftentimes challenging to directly translate these experimental results into implications of messy human behaviors in the past due to multiple reasons.

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as their experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in both philosophy of science [@cartwright2007] and econometrics [@deaton2018].

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience [@shamay-tsoory2019; @sonkusare2019], a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment. Unlike controlled experiments, variability could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in recent years [@cunningham2021; @hernan2023]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

# The ethology and ethnography of stone toolmaking

As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentionality simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999]. The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].

@@

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner, while novices often feature a low planning depth [@vanopheusden2023] and a different set of perception on the causal structure of how certain behaviors will modify the raw materials.

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)). Here we use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019]. The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1). Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.

@@

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022].

Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities. This approach enables researchers to gain a deeper understanding of how people in ancient societies might have perceived and interpreted their environment, objects, and actions. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.
.

# The curse of knowledge

Contemporary practices in experimental archaeology, as manifested by the fact that the majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher, tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as grandiose generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

It is essential to recognize that modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years. This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders. By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts. This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise. It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

# Open science beyond reproducibility

The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, while videos are the main vector of process-level data, featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

The accessibility and availability of experimental data foster collaboration and enhance the reproducibility and transparency of research findings, as others can verify and validate the results by examining the original data. More importantly, a centralized database also promotes data preservation and long-term accessibility. By storing experimental data in a structured and organized manner, it safeguards valuable information from potential loss or degradation over time. This preservation ensures that the data remains accessible for future researchers, avoiding the loss of valuable insights and preventing the need for redundant and costly repetitions of experiments. It also allows for the reanalysis of existing data, facilitating discoveries and insights that may not have been initially anticipated. However, it has been widely acknowledged that the reuse of archaeological data has not received enough attention among researchers in our discipline [@faniel2018; @huggett2018; @moody2021]. Among many reasons preventing archaeologists from reusing published and digitized data [@sobotkova2018], the lack of a standardized practice of and motivation for data sharing is a prominent one [@marwick2018].

@@

This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach" (C. L.).

>>>>

---
title: "Expanding the scope of experimental archaeology using the Perception-Process-Product conceptual framework"
author: 
- Cheng Liu^[Department of Anthropology, Emory University, Atlanta, GA, USA; raylc1996@outlook.com]
abstract: "This paper presents the outline of the Perception-Process-Product ('Triple P') conceptual framework that aims to expand the scope of experimental archaeology. The Triple P framework emphasizes the amplification of multi-level variabilty and the identification of causal relationships of variations across the levels of perception, process, and product. Here I propose the following five basic measures to put the Triple P framework into practice: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects.  \\par \\textbf{Keywords:} Experimental archaeology; Ethological analysis; Ethnographical analysis; The curse of knowledge; Collaborative knowledge production"
date: "`r Sys.Date()`"
output:
  word_document
linestretch: 1.5
link-citations: yes
linkcolor: blue
bibliography: bibliography.bib
csl: apa.csl
fontsize: 11pt
always_allow_html: true
---

# Introduction

This paper presents the Perception-Process-Product (hereinafter referred to as "Triple P") conceptual framework to expand the scope of experimental archaeology. Adopting the principle of Occam's razor [e.g., @domínguez-rodrigo2008; @reeves2009], whether explicitly or implicitly, the field of experimental archaeology has long tended to center around the reverse engineering of a past technology in a minimal or least-effort manner while ignoring the rich contextual information it affords. Nevertheless, Occam's razor, this very cornerstone of our field, can be insufficient to infer the preferences of irrational agents possessing incomplete information [@armstrong2018] in tool design and use, and the two conditions described here, I believe, provide a better approximation of past humans displaying extensive cultural variability as opposed to the assumption of omniscient *Homo economicus* for most social scientists in the post-Kahneman era, but particularly anthropologists and archaeologists [@henrichSearchHomoEconomicus2001].

As such, built upon early intellectual principles and practices in behavioral archaeology [@schiffer2010], here I propose the Triple P framework, which aims to **a)** amplify the expression of variability in experimental replicas (product) and their associated behavioral channels (process) as well as sensory experiences (perception) and **b)** better identify the complex interacting relationships across these three levels of variations. To accomplish these two objectives, I advocate the following three measures as integral components of the Triple P framework: 1) acknowledging the contribution and limitations of actualistic experiments properly; 2) encouraging collaborative projects that involves geographically diverse and non-traditional research participants such as hobbyists and novices; 3) adopting a workflow that normalize the collection and curation of ethological and ethnographic data in experimental projects. It is no doubt that strategies of data collection and analysis of a given experimental project should be primarily derived from the research question, which can be legitimately narrow in scope, but the awareness of the rich toolkit available can sometimes inspire researchers to ask questions that are bold and transformative [@schmidt2020]. Here I will mainly leverage the extensive corpus in experimental designs and inferences revolving around stone artifacts to clarify its meaning and demonstrate the necessity and potential of this framework.

@@

The trade-off between causality (aka "internal validity") and generalizability (aka "external validity") forms a central issue in experimental design [@eren2016; @roe2009: 1266-1267], and the past decades have witnessed experimental archaeology's growing research interests focusing on the causal mechanism at the behavioral level in the explanation of material culture variation [@eren2016; @lin2018]. In the context of stone artifact replication, one typical research design emphasizing causality over generalizability is the use of knapping machines/robots [@li2022; @pfleging2019], which has helped map out the physical constraints of stone artifact manufacture and use through the identification of multiple groups of causal relationships between input (force, exterior platform angle, platform depth, etc.) and outcome variables (flake size, flake shape, wear formation, etc.). All variables of interest in this setting are relatively easy to measure, quantify, and control, but this type of design is incapable of inferring how context-generic principles interact in a particular context as reflected in real-world conditions. In addition to the applications of machine knapping, the same problem is also incurred by the introduction of standardized artificial material like bricks [@lombao2017] or foam blocks [@schillinger2016] in experimental studies focusing on the transmission of lithic technologies, which was demonstrated to be problematic [@liu2023]. In reality, each rock has a different shape and often different physical properties such as inner cracks and inclusions, and this heterogeneity itself represents a critical variable in the process of skill acquisition. After all, these experimental results can only be as robust as their experimental settings.

More importantly, controlled experiments without randomization are not enough to infer causal mechanisms, which may be severely biased by factors such as individual differences [@pargeter2023], allocation concealment [@schulz2002], and poor recruitment [@fletcher2012]. This issue is rarely discussed in material science oriented experiments but should raise a critical concern in experimental archaeology as social science. Randomized Controlled Trials (RCT) *sensu stricto* as practiced in contemporary medicine and behavioral sciences, known for its high cost [e.g., @speich2019], are extremely rare in experimental archaeology when human participants were involved. Rather, most of our knowledge regarding the past is derived from data sets that can be characterized as Small, Unbalanced, Noisy, but Genuine (SUNG) [@arnaud2023] produced through experiments featuring small-sized convenience sample. It has also been a debatable issue whether Randomized Controlled Trial (RCT) represents the gold standard of knowledge in the realm of philosophy of science [@cartwright2007].

On the other hand, actualistic experiments pay more attention to how experimental insights can be generalized to archaeological samples by incorporating authentic materials and plausible social settings with a certain degree of compromised control [@outram2008]. Back to the cases of cultural transmission experiments, an actualistic experiment would involve the use of natural rocks with varied morphology instead of standardized artificial materials as well as human demonstrators instead of videos of knapping instruction, despite the fact that the latter will remain consistent across individuals. Interestingly, researchers in cognitive psychology and neuroscience, a field known for its development of rigorous and well-controlled experimental methods, have started to use naturalistic stimuli more frequently and advocate a paradigm shift to semi-controlled experiment [@shamay-tsoory2019; @sonkusare2019; @nastase2020; @yarkoni2022]. Unlike controlled experiments, variability could be easily observed in actualistic experiments by design. This feature is crucial and cannot be simply replaced by ethnographic records, because many paleolithic technological components are not displayed in contemporary non-industrial societies, which usually feature technological systems with groundstone artifacts as the target products [e.g., @stout2002; @arthur2018]. Furthermore, statistical techniques for developing causal inference from observational data, which essentially represent the nature of results from actualistic experiments, have also been greatly boosted in epdedimiology and economics in recent years [@cunningham2021; @hernan2023; @nichols2007]. Lastly, actualistic experiment can serve as a heuristic for hypothesis generation, aligning with the perspective of Lin et al. [-@lin2018: 680-681] and the argument put forth by Ingersoll and MacDonald [-@ingersoll1977], who proposed that the interaction between actualistic and controlled experiment "operates in a cyclical form of induction and deduction."

Contemporary practices in experimental archaeology, as manifested by the fact that a majority of scholarly publications are produced as results of experiments conducted by a single knapper with a dual identity of researcher [@whittaker1999], tend to be restrained by the cognitive bias known as the "curse of knowledge" or "curse of expertise". The curse of knowledge refers to the phenomenon that it is extremely challenging for experts to ignore the information that is held by them but not others, particularly novices [@hinds1999; @camerer1989]. When the knapping expertise is gradually formed through multiple years of observations and trial-and-error learning, an expert knapper develops some specific ways of strategic planning, motor habits (and their associated impacts on anatomical forms like wrist and elbow), preferences of percussor and raw material types, as well as familiarity of various techniques that become unforgettable [@moore2020: 654]. The existence of this cognitive bias is not inherently bad, and these many years of experiences should be appreciated and celebrated by experimental archaeologists. However, what is problematic is that the results of replication experiments conducted by these experienced practitioners, often in settings of single knapper, has been constantly framed as generalization regarding the evolution of technology and cognition that masks a huge range of technological diversity.

Modern flintknapping techniques, as a research subject and a scientific method, originated from hobbyists' individualistic trials of reverse engineering during the 19th century, rather than from the inter-generational transmission of knapping knowledge spanning millions of years. This historical context is well elucidated by studies on the subject [@flenniken1984; @reevesflores2010; @coles1979; @johnson1978; @whittaker1994: 54-61]. Hobbyist knappers represent a huge repertoire of technological knowledge that does not fully overlap with what is acquired by academic knappers. They tend to come up with ideas that may appear to be counter-intuitive at first glance for academics. One such example is the utility of obtuse edge angle as demonstrated by Don Crabtree[-@crabtree1977], a mostly self-educated flintknapper yet one of the most important figures in experimental archaeology. In his experiment, Crabtree demonstrated the excellent performance of blade dorsal ridge on tasks like shaving and cutting hard materials, challenging the traditional perspective on producing sharp lateral edges as the sole purpose of stone toolmaking and shedding light on future functional reconstruction through the use-wear analysis. It is rather unfortunate that collaborations between academics and hobbyists are less common than expected due to their complicated and uneasy relationships as detailed in Whittaker's [-@whittaker2004] famous ethnography. Likewise, novices' lack of expertise also helps to mitigate the "curse of knowledge" bias that may hinder expert knappers. Their involvement can potentially lead to the discovery of alternative methods, techniques, and interpretations that may have been overlooked by experts.

Emphasizing variability at its core, the Triple P conceptual framework inherently adopts a collaborative mode of knowledge production, which has been recently advocated in experimental studies [@ranhorn2020; @liuInferringCulturalReproduction2023] and museum collection studies [@timbrell2022] of stone artifacts. The Triple P framework recognizes that experimental archaeology can greatly benefit from diverse perspectives and contributions from multiple stakeholders. By engaging researchers, practitioners, and local communities from different geographical locations, the framework acknowledges the importance of including voices from various cultural backgrounds and contexts [@pargeter2023: 164]. This emphasis on collaboration and inclusivity allows for a more nuanced understanding of the complexities of raw material procurement [@batalla2016], selection [@arthur2021], pre-treatment [@maloney2020], production [@griffin2013], and use [@martellotta2022] across different regions. Furthermore, the Triple P framework promotes the recognition and value of local knowledge and expertise. It acknowledges that communities living in specific geographical areas possess unique insights and understanding of their cultural heritage. By involving these local communities in the research process, the framework allows their voices to be heard and their contributions to be acknowledged. This not only enhances the quality of research outcomes but also fosters a sense of ownership and pride within these communities, strengthening the connection between archaeological research and the people it directly affects [@douglass2020; @marshall2002].

# The Triple P framework in action

As implied in its name, the implementation of the Triple P framework involves the collection of process-level (ethological) and perception-level (ethnographic) data (**Figure** \@ref(fig:concept)), which is critical to address equifinality and multifinality [@hiscock2004; @premoEquifinalityExplanationThoughts2010; @nami2010], two daunting challenges in archaeological inference that partially contributed to the discipline-wide paradigm shift in the 1980s [@lake2014: 264-265]. Equifinality refers to the phenomenon where a similar state or consequence can be achieved through multiple different paths, while multifinality emerges when a similar process can lead to multiple ends. While we cannot fully solve these two problems and accurately reconstruct the past behavioral processes and intentions simply based on materials remains, context-rich experiments involving the collection of ethological and ethnographic data can help us better document an enlarged range of possible combinations of variation at these three levels and thereby evaluate the probability of certain behavioral mechanisms behind a given archaeological assemblage [@reynolds1999; @stout2023]. The importance of specifying and documenting the context information of both the experiment as well as the phenomenon of interest has also been recently highlighted in psychological sciences [@holleman2020].

@@

## Product-level data

Traditionally speaking, the product-level data, namely the documentation and analysis of replicas, form the sole research subject of experimental archaeology and serve as the tangible foundation for analogical inference in the interpretation of archaeological materials. It can exist in the form of spreadsheets containing detailed technological attributes, or photos and illustrations, or high-resolution 3D scans of individual artifact or a whole assemblage. No particular modification regarding the collection procedure of product-level data is required in the context of the Triple P framework, although it should be emphasized that the definition of variables measured and the documentation techniques (models of camera/scanners, light setting, processing software version and workflow, etc.) should be always available in the relevant meta-data. Besides, adopting good habits in spreadsheet data organization is strongly recommended [@broman2018].

## Process-level data

While formal ethological methods that are widely used in the description and analysis of non-human animal behavior [@fragaszy2018] still largely fall into oblivion among archaeologists, the attempts of reconstructing behavioral sequences involved in the manufacture of material remains are not infrequent. One such example is cognigram, which was first systematically developed and applied in the archaeological research by Haidle [@haidleHowThinkSimple2009; @haidle2010; @lombard2012; @haidle2023]. Cognigram essentially represents an abstracting process of a series of action sequences achieving a similar goal. This approach is a powerful and elegant yet limited by the curse of expertise [@hinds1999], meaning it cannot handle variability very well. To some extent, it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assumes clear causal thinking between each step in an idealistic manner. However, this may be biased given that 1) experiments and modeling studies in cognitive science have shown that novices often feature a low planning depth [@vanopheusden2023] and 2) ethnographic studies demonstrated that even expert practitioners in traditional societies can have a different set of perception on the causal structure of how certain behaviors will modify the raw materials [@harris2021].

Consequently, we need to accumulate more real-world data by recording a large amount videos of toolmaking and conducting systematical ethogram analysis. With the emergence of new software platforms such as BORIS [@friard2016], the difficulty of coding has decreased significantly in recent years (**Figure** \@ref(fig:ethogram)). Here I use a modified version of action grammar developed by [@stout2021] as an example, among multiple coding schemes featuring different research focus [@muller2023]or degrees of coding complexity [@mahaney2014; @cueva-temprana2019]. The knapping action recorded in videos can be coded following the ethogram presented in **Table** \@ref(tab:tab1). Depending on the original research question, sequences of coded actions can then be used in further analysis, such as complexity [@stout2021], similarity [@mobbs2021], etc.

@@

## Perception-level data

Ethnographies revolving around general archaeological practices [@ethnogra2006], experimental archaeology as a field [@reevesflores2012], as well as practices of specific technologies like flintknapping, including both WEIRD [@whittaker2004] and non-WEIRD populations[@stout2002; @arthur2018], are far from novel. However, it has never been formally recognized as a legitimate research method in experimental archaeology. Echoing with the recent trends of adopting embodied cognition [@varela2017] in archaeological research [@malafouris2013], ethnographic data and methods can reveal hidden information that is otherwise irretrievable and thus should occupy a unique niche in experimental archaeology. This also echoes the post-positivist turn in psychology, a field that is known for the development of experimental methods, in the past decades, particularly the emphasis on the value of incorporating qualitative research [@syed2022; @weger2019; @stoutCognitiveScienceTechnology2021].

Through participant observation, interviews, and detailed field notes, ethnographers can capture the subtle nuances of perception, such as sensory experiences, social interactions, and cultural meanings associated with the experimental activities. Compared with the ethological methods, the interview questions and participant observation in ethnographic methods feature an even higher degree of freedom and rely more heavily on the research question as well as ad-hoc interaction. One potential application of ethnographic methods in experimental archaeology of stone artifacts is asking knappers about the intentions of each action and see how it matches with the results as revealed by lithic analysis of replicas, which can provide crucial contextual information addressing the issues of equifinality and multifinality in the formation of lithic assemblage.

## Multi-level data curation

The comparative study and large-scale synthesis of variability data require the building of centralized, open-access, and carefully curated data infrastructure [@marwick2017], which unfortunately still does not exist yet in experimental archaeology and likely won't be available in the near future. Among the three dimensions of the Tripe P framework, the product-level data are usually stored in the format of spreadsheets, photos, and 3D models, and the perception-level data formats mainly include audio files and their transcribed texts, whereas videos are the main vector of process-level data, a rather non-traditional data format in archaeological research featuring the highest file size compared with the other two. As such, following data sharing principles of FAIR [@wilkinson2016] and CARE [@carroll2020], the Triple P framework recommends Databrary [@simon2015; @gilmore2015], a web-based library that was originally designed for developmental scientists, as the main data curation platform, where researchers can freely upload video files and related metadata that can connect with different types of data within the same project.

@@

I would like to thank Dietrich Stout for his helpful comments on earlier drafts of this article. I am also grateful for my collaborators who help me turn this idea into reality: Russell Cutts (Oxford College of Emory University), Nada Khreisheh (The Ancient Technology Centre), Justin Pargeter (New York University), Ron Shimelmitz (University of Haifa), and particularly Mark Moore (University of New England). This study was supported by a research grant from the Leakey Foundation titled "Inferring skill reproduction from stone artifacts: A middle‐range approach."
